The Economics of Wildfire Aviation

### 2023-10-28 manuscript/manuscript.qmd
almost final revisions

Aviation contracts are complex legal documents, but generally include a number of safety requirements, a daily availability price, and an hourly flight price. The two forms of contract are Exclusive Use (EXU) and Call-When-Needed (CWN). Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at a daily cost. Call-when-needed contracts are more flexible and can be activated on short notice [@thompson2012]. However, CWN contacts generally specify a daily availability rate and hourly flight rate that is higher than EXU contracts. The question is: how many of each contract should the USFS sign at the start of each fire season?[^2] The answer depends on expected use patterns of the aviation resources. We focus on this question in the context of large and expensive aircraft known as Large Airtankers (LAT) and Very Large Airtankers (VLAT) classified based on their water and retardant capacity.

@@

Cost benefit analysis requires the accounting of the costs and benefits of an intervention over a relevant time period [@boardman2018]. The financial cost of using aviation to fight fire is recorded because the government is responsible for payment.[^4] The benefits of aviation are harder to quantify because they involve the avoided costs and losses of fires that were prevented from growing and damaging valued assets. If aviation resources reduce the probability that a fire spreads and damages property, we do not observe the counterfactual - what would have happened in the absence of aviation use. This section describes how aviation resources are currently used to fight fire, discusses the conceptual challenges to measuring benefits, and proposes data driven approaches to measuring the benefits of aviation given currently available data.

@@

[^7]: Assets like homes generally have observable values, whereas cultural heritage sites may not. Moreover, many object to the strict use of monetary value of assets that may imply the benefits of protecting high-value homes are higher than protecting low-value homes. The principles described here accommodate alternative definitions of values at risk that may include subjective values.

>>>>

Aviation contracts are complex legal documents, but generally include a number of safety requirements, a daily availability price, and an hourly flight price. The two forms of contract are Exclusive Use (EXU) and Call-When-Needed (CWN). Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at a daily cost. Call-when-needed contracts are more flexible and can be activated on short notice [@thompson2012]. However, CWN contacts generally specify a daily availability rate and hourly flight rate that is higher than EXU contracts. The question is: how many of each contract should the USFS sign at the start of each fire season?[^2] The answer depends on expected use patterns of the aviation resources. We focus on this question in the context of large and expensive aircraft known as Large Airtankers (LAT) and Very Large Airtankers (VLAT) classified based on their water and retardant capacity. However, our general approach could be applied to other aviation resources as well.

@@

Cost benefit analysis requires the accounting of the costs and benefits of an intervention over a relevant time period [@boardman2018]. The financial cost of using aviation to fight fire is recorded because the government is responsible for payment.[^4] The benefits of aviation are broad, including the deployment of retardant for direct suppression, monitoring the fire and resources, and transporting people and equipment. We focus on the benefits of direct suppression in this analysis. These benefits are hard to quantify because they involve the avoided costs and losses of fires that were prevented from growing and damaging valued assets. If aviation resources reduce the probability that a fire spreads and damages property, we do not observe the counterfactual - what would have happened in the absence of aviation use. This section describes how aviation resources are currently used to fight fire, discusses the conceptual challenges to measuring benefits, and proposes data driven approaches to measuring the benefits of aviation given currently available data.

@@

[^7]: We use the term asset to include private assets like homes and property as well as public assets such as ecosystem service values. The value of homes is generally well-understood because transaction prices are observable. Ecosystem service values must be measured via non-market valuation, and are generally not salient to incident management teams during a fire event [@venn2011].
